CrystalPlan: an Experiment Planning Tool for Crystallography

### 2010-09-21 docs/CrystalPlan_paper/paper.tex
Refs #182: bibliography and figures set up.

     %-------------------------------------------------------------------------
     % Information about the type of paper
     %-------------------------------------------------------------------------
     \paperprodcode{a000000}      % Replace with production code if known
     \paperref{xx9999}            % Replace xx9999 with reference code if known
     \papertype{FA}               % Indicate type of article
                                  %   FA - research papers (full article)
                                  %   SC - short communications
                                  %   LA - lead article
                                  %   FE - feature articles
                                  %   ST - structural communications
                                  %   XC - crystallization communications
                                  % (Following categories rarely in LaTeX)
                                  %   AA - abstracts
                                  %   AD - addenda and errata
                                  %   BC - books received
                                  %   BR - book reviews
                                  %   CA - cif applications
                                  %   CE - current events
                                  %   CI - inorganic compounds
                                  %   CM - metal-organic compounds
                                  %   CN - cryocrystallography papers
                                  %   CO - organic compounds
                                  %   CP - computer programs
                                  %   CR - crystallographers
                                  %   CS - scientific comment
                                  %   ED - editorial
                                  %   EI - inorganic compounds
                                  %   EM - metal-organic compounds
                                  %   EO - organic compounds
                                  %   FI - inorganic compounds
                                  %   FM - metal-organic compounds
                                  %   FO - organic compounds
                                  %   IP - issue preface
                                  %   IU - iucr
                                  %   LE - letters to the editor
                                  %   LN - laboratory notes
                                  %   ME - forthcoming meetings/short courses
                                  %   MR - meeting reports
                                  %   NN - notes and news
                                  %   NP - new commercial products
                                  %   OB - obituaries
                                  %   SR - software reviews
                                  %   TE - teaching and education

@@

     % The title of the paper. Use \shorttitle to indicate an abbreviated title
     % for use in running heads (you will need to uncomment it).

\title{CrystalPlan: an Experiment Planning Tool for Crystallography}
\shorttitle{CrystalPlan: an Experiment Planning Tool}

@@

     % Keywords (required for Journal of Synchrotron Radiation only)
     % Use the \keyword macro for each word or phrase, e.g. 
     % \keyword{X-ray diffraction}\keyword{muscle}

%\keyword{keyword}

@@

%-------------------------------------------------------------------------
%-------------------------------------------------------------------------
%---------------------------------------------------------------------------------------
\section{Introduction}

@@

CrystalPlan is designed to be able to use a variety of sample orientation
goniometers. A base Goniometer class is subclassed for specific types of
goniometers, allowing a unified API: in a way, it is a “universal goniometer”,
which uses the conventional goniometer angles of phi, chi, omega as defined in
[REF XXX]. Each class inheriting from Goniometer can convert the desired phi,
chi, omega angles to the internal motor positions it requires. The API allows
for limited goniometers: for example, TOPAZ is currently equipped with a
goniometer where chi is fixed at 45 degrees. CrystalPlan handles this by
limiting the available sample orientation angles accessible to the user.

@@

For each sample orientation, a rotation matrix combining the sample's mounting
orientation (the U matrix loaded from ISAW) and the goniometer rotation
($\Phi$, $\Xi$, $\Omega$) is computed. The coverage in reciprocal space is
computed in this way: for each pixel in the face of a detector, the direction of
the scattered beam is known. Then, two incident beam vectors are considered,
corresponding to the low and high wavelength limits of the detectors and/or
source. The scattered beam direction and the two possible incident beam vectors
are used to to find two q vectors, which point in the same direction. At the
high wavelength we find qmin, and the low wavelength limit scattering off of qmax.
Since the instrument has a wide bandwidth, we fill in the 3D q-space matrix by
filling all points from qmin to the  qmax, for all pixels in the face of the
detector. This quickly defines a volume of coverage, as shown in Figure X.

@@

Crystal lattice parameters (lengths a,b,c and angles alpha, beta, gamma) are
typed in or loaded from the ISAW UB matrix file. CrystalPlan then generates a
list of HKL peaks with d-spacings larger than the $d_{min}$ specified. For each
sample orientation being simulated and for each peak, a function calculates at
what wavelength lambda and in which direction it scatters. If the wavelength is
within range of the limits of the instrument, the beam direction is projected
onto each detector face to determine where, on each detector surface, it will be
measured.

@@

When entering the lattice parameters, the crystal's point group symmetry can
also be entered. This information is then used to calculate the increased
measurement redundancy. A list of n 3x3 multiplication matrices is generated
based on the point group; any {hkl} vector can be multiplied by these matrices
to find the n equivalent hkl reflections; for example, for orthorombic (mmm)
symmetry, n=8 and 8 multiplication matrices are generated, which multiply {hkl}
to give {±h, ±k, ±l}.

@@

CrystalPlan allows the user to manually enter lists of sample orientations to
simulate; however, finding an optimal plan by trial and error would be very time
consuming. To make the process easier, CrystalPlan includes an Automatic
Coverage Optimizer, which takes as its inputs the desired number of sample
orientations and uses a genetic algorithm to find a solution that satisfies a
user-specified coverage criterion (for example, the user may ask for 85% of
peaks to be measured, or for 90% of q-volume).

@@

\subsubsection{Main Program Workflow}
Some screenshots of the GUI are shown in Figure Y. The main window consists of a
series of tabs, with workflow proceeding from left to right:

@@

\subsubsection{Reciprocal Space Coverage Visualization}
Figures Z -a shows a screenshot of the reciprocal space coverage 3D interface.
The 3D view can represent q-space in two ways: volume coverage, or single-peak
coverage.

@@

The reflections view (Figure ZZ-a) displays each hkl as a small sphere; the
color of which indicates the number of times the hkl is predicted to be measured
in the current experiment plan.

Both views can take into crystal symmetry in their display (Figures Z-b and
ZZ-b), using the techniques described previously.

@@

     % References are at the end of the document, between \begin{references}
     % and \end{references} tags. Each reference is in a \reference entry.

\begin{references}
\reference{Author, A. \& Author, B. (1984). \emph{Journal} \textbf{Vol}, 
first page--last page.}
\end{references}

@@

\begin{figure}
\caption{Caption describing figure.}
%\includegraphics{fig1.ps}
\end{figure}

>>>>

     %-------------------------------------------------------------------------
     % Information about the type of paper
     %-------------------------------------------------------------------------
     \paperprodcode{a000000}      % Replace with production code if known
     \paperref{xx9999}            % Replace xx9999 with reference code if known
     \papertype{FA}               % Indicate type of article
                                  %   FA - research papers (full article)
                                  %   SC - short communications
                                  %   LA - lead article
                                  %   FE - feature articles
                                  %   ST - structural communications
                                  %   XC - crystallization communications

@@

\title{CrystalPlan: an Experiment Planning Tool for Crystallography}

\shorttitle{CrystalPlan: an Experiment Planning Tool}

@@

@@

%-------------------------------------------------------------------------
%-------------------------------------------------------------------------
%-------------------------------------------------------------------------
\section{Introduction}

@@

CrystalPlan is designed to be able to use a variety of sample orientation
goniometers. A base Goniometer class is subclassed for specific types of
goniometers, allowing a unified API: in a way, it is a “universal goniometer”,
which uses the conventional goniometer angles of $\phi$, $\chi$, $\omega$ as
defined in \cite{knuth84}. Each class inheriting from Goniometer can convert
the desired phi, chi, omega angles to the internal motor positions it requires. The API allows
for limited goniometers: for example, TOPAZ is currently equipped with a
goniometer where chi is fixed at 45 degrees. CrystalPlan handles this by
limiting the available sample orientation angles accessible to the user.

@@

For each sample orientation, a rotation matrix combining the sample's mounting
orientation (the U matrix loaded from ISAW) and the goniometer rotation
($\Phi$, $\Xi$, $\Omega$) is computed. The coverage in reciprocal space is
computed in this way: for each pixel in the face of a detector, the direction of
the scattered beam is known. Then, two incident beam vectors are considered,
corresponding to the low and high wavelength limits of the detectors and/or
source. The scattered beam direction and the two possible incident beam vectors
are used to to find two q vectors, which point in the same direction. At the
high wavelength we find qmin, and the low wavelength limit scattering off of qmax.
Since the instrument has a wide bandwidth, we fill in the 3D q-space matrix by
filling all points from qmin to the  qmax, for all pixels in the face of the
detector. This quickly defines a volume of coverage, as shown in Figure
\ref{fig:volume_coverage}.

@@

Crystal lattice parameters (lengths $a$, $b$, $c$ and angles $\alpha$, $\beta$,
$\gamma$) are typed in or loaded from the ISAW UB matrix file. CrystalPlan then
generates a list of HKL peaks with d-spacings larger than the $d_{min}$ specified. For each
sample orientation being simulated and for each peak, a function calculates at
what wavelength lambda and in which direction it scatters. If the wavelength is
within range of the limits of the instrument, the beam direction is projected
onto each detector face to determine where, on each detector surface, it will be
measured.

@@

When entering the lattice parameters, the crystal's point group symmetry can
also be entered. This information is then used to calculate the increased
measurement redundancy. A list of $n$ $3\times3$ multiplication matrices is
generated based on the point group; any {hkl} vector can be multiplied by these matrices
to find the n equivalent hkl reflections; for example, for orthorombic (mmm)
symmetry, $n=8$ and 8 multiplication matrices are generated, which multiply
{hkl} to give {±h, ±k, ±l}.

@@

CrystalPlan allows the user to manually enter lists of sample orientations to
simulate; however, finding an optimal plan by trial and error would be very time
consuming. To make the process easier, CrystalPlan includes an Automatic
Coverage Optimizer, which takes as its inputs the desired number of sample
orientations and uses a genetic algorithm to find a solution that satisfies a
user-specified coverage criterion (for example, the user may ask for $85\%$ of
peaks to be measured, or for $90\%$ of q-volume).

@@

\subsubsection{Main Program Workflow}
Some screenshots of the GUI are shown in Figure \ref{fig:screenshots}. The main
window consists of a series of tabs, with workflow proceeding from left to right:

@@

\subsubsection{Reciprocal Space Coverage Visualization}
Figure \ref{fig:volume_view}a) shows a screenshot of the reciprocal space
coverage 3D interface. The 3D view can represent q-space in two ways: volume coverage, or single-peak
coverage.

@@

The reflections view (Figure \ref{fig:refl_view}a) displays each hkl as a
small sphere; the color of which indicates the number of times the hkl is predicted to be measured
in the current experiment plan.

Both views can take into crystal symmetry in their display (Figures
\ref{fig:volume_view}b and \ref{fig:refl_view}b), using the techniques described
previously.

@@

     % References are at the end of the document, between \begin{references}
     % and \end{references} tags. Each reference is in a \reference entry.
\bibliographystyle{iucr}     
\bibliography{crystalplan}

%\begin{references}
%\reference{Author, A. \& Author, B. (1984). \emph{Journal} \textbf{Vol}, 
%first page--last page.}
%\end{references}

@@

\begin{figure}
\caption{volume coverage}
%\includegraphics{fig1.ps}
\label{fig:volume_coverage}
\end{figure}

\
\begin{figure}
\caption{Screenshots of the main GUI}
%\includegraphics{fig1.ps}
\label{fig:screenshots}
\end{figure}

\begin{figure}
\caption{Volume coverage in reciprocal space view: a) without crystal symmetry;
b) with crystal symmetry}
%\includegraphics{fig1.ps}
\label{fig:volume_view}
\end{figure}

\begin{figure}
\caption{Coverage of single-crystal reflections in reciprocal space: a) without
crystal symmetry; b) with crystal symmetry}
%\includegraphics{fig1.ps}
\label{fig:refl_view}
\end{figure}

\begin{figure}
\caption{figureeeeee}
%\includegraphics{fig1.ps}
\label{fig:screenshots}
\end{figure}
